How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-11-05 doc/article.tex
pretty head

While all of our models are conditioned on past extinction data from the Cenozoic planktonic microfossil record, we used multiple ways to encode historical information which describes the species over time such as change in geographic range or previous climate state or by allowing covariate effects to vary over time. Models that include the historical covariates (e.g. change in geographic range) but do not allow covariate effects to vary over time (i.e. CP model) encode the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, as with our V model, does not explicitly encode the history of individual species into our model but instead models the history of how covariate effects vary over time, thus implicitly encoding historical information about the species in our models. By modeling the variation in covariate effects over time, forecasts made for future extinction events are conditioned on a wider range of potential covariate effects which can improve our models flexibility when forecasting extinction in the novel environmental conditions we might expect in the future. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}). Our results complement those of \citet{Kiessling2016} who found that including historical range trajectories improved the performance of extinction risk forecasts.

The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many Foraminifera species \citep{Ezard2011}. However, comparable information does not exist for all Foraminifera species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the Foraminifera may be informative.

@@

To summarize, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors. Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models. The remaining three models yield nearly equivalent out-of-sample forecasts. The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data, and that historical information. Additionally, including historical information via explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.

@@

We thank Jonathan Payne, Thomas Ezard, Erin Saup, and Peter Roopnarine for useful discussions. Wolfgang Kiessling, Matthew Clapham, and an anonymous reviewer provided thoughful reviews that strengthened the manuscript consideraly. This work was funded by a David and Lucile Packard Fellowship to SF.

>>>>

While all of our models are conditioned on past extinction data from the Cenozoic planktonic microfossil record, we used multiple approaches to encode historical information. Models that include the historical covariates (e.g. change in geographic range) but do not allow covariate effects to vary over time (i.e. CP model) encode the past explicitly but assumes that covariates effects are constant over time. Allowing covariate effects to vary over time, as with our V model, does not explicitly encode the history of individual species into our model but instead models the history of how covariate effects have varied over time, thus implicitly encoding historical information about the species in our models. By modeling the variation in covariate effects over time, forecasts made for future extinction events are conditioned on a wider range of potential covariate effects which can improve our models flexibility when forecasting extinction in the novel environmental conditions we might expect in the future. Comparing our out-of-sample forecast results indicates that these approaches yield approximately equal forecasting performance (Fig. \ref{fig:fold_auc}). Our results contrast somewhat with those of \citet{Kiessling2016}, who found that including historical range trajectories significantly and substantially improved the performance of extinction risk forecasts. Whether this reflects differences in analytical methodology or study system (planktonic microfossil taxa versus marine macroinvertebrates) is not clear and is worth further investigation.

The relative quality and consistency between in-sample and out-of-sample forecasting performance for three of the four models we considered is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many Foraminifera species \citep{Ezard2011}. However, comparable information does not exist for all Foraminifera species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the Foraminifera may be informative.

@@

To summarize, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of simple taxonomic, geographic, and historical predictors. Although a model that includes historical covariates such as change in geographic range and change in climate between observations while also allowing covariate effects to vary over time performs best at in-sample prediction, this model is overfit to our data and produces less accurate out-of-sample forecasts than our three less complex models. The remaining three models yield nearly equivalent out-of-sample forecasts. The results of this simple exercise suggest that conservation decisions could indeed be bolstered by including fossil data. Additionally, including historical information via explicit modeling of historical covariate effects or modeling how covariate effects have changed over time does not diminish and may ultimately improve our ability to forecast future extinctions.

@@

We thank Jonathan Payne, Thomas Ezard, Erin Saupe, and Peter Roopnarine for useful discussions. Wolfgang Kiessling, Matthew Clapham, and an anonymous reviewer provided thoughtful reviews that strengthened the manuscript considerably. This work was funded by a David and Lucile Packard Fellowship to SF.
